Risk Factors and Treatment of Refractory Anaphylaxis - a review of case reports

### 2018-02-23 analysis/paper/paper.Rmd
changes post 1st rev round + Prof Worm

Department of Dermatology, Venerology, and Allergology, Charité – Universitätsmedizin Berlin, corporate member of Freie Universität Berlin, Humboldt-Universität zu Berlin, and Berlin Institute of Health, Charitéplatz 1, 10117 Berlin,

@@

# Introduction
Anaphylaxis is a potentially life-threatening, primarily mast cell-dependent reaction to most frequently food, drugs, insect venom. Although anaphylaxis remains to be a clinical diagnosis [@Kemp2014], it is highly likely if any of the following 3 criterita are fulfilled: 1) Acute onset of an illness with generalized skin / mucosal involvemt with flushing, hives or pruritus; 2) acute cutaneous, cardiorespiratory or gastrointestinal symptoms after exposure to a likely allegen; c) acute reduction of blood pressure (BP) after exposure to a known allergen [@Sampson2006]. The definition of anaphylaxis remains controverse [@Lieberman2015]. The pathogenesis of anaphylaxis is related to a sudden, systemic, (mostly IgE dependent) degranulation of mast cells or basophils leading to the release of multiple mediators like histamine, tryptase, chymase, platelet-activating factor, prostaglandins, and leukotrienes [@Houser2015].

By contrast, the definition of refractory anaphylaxis has not been established yet. For the purpose of this review, we define refractory anaphylaxis as anaphylaxis (meeting the criteria by NIAID/FAAN  [@Sampson2006]) unresponsive to the treatment with at least 2 doses of a minimum 300 µg adrenaline. Unresponsiveness, in this case, is defined as a lack of expected normalization of clinical symptoms (i.e. rapid increase in blood pressure, bronchodilatation, tachycardia). Previously authors suggested using the term in cases of persistent severe shock symptoms after 10 minutes of adequate resustitation [@Gouel2016].

It is important to distinguish refractory anaphylaxis (which is unresponsive to treatment) from recurrent anaphylaxis which can be defined as multiple episodes of anaphylaxis over a period of time that responsed to treatment with adrenaline [@Mullins2003]. In one case report authors used the term refractory anaphylaxis to describe 6 recurrent episodes of anaphylaxis associated with the monoclonal mast cell activation syndrome, which responded well to intramuscular doses of epinephrine [@Jagdis2014] and were ultimately treated with omalizumab. We did not include such reports into our analysis and suggest to use the term “recurrent” anaphylaxis instead of “refractory” in such cases.

The incidence of anaphylaxis is increasing worldwide [@Koplin2011] and was lately estimated to be between 50 and 112 episodes per 100,000 person-years and an estimated prevalence of 0.3% to 5.1% [@TejedorAlonso2015].  It is widely believed to be an underrecognized and undertreated medical emergency [@Fineman2015]. Anaphylaxis in children is in the majority of cases food dependent and most likely less severe than episodes in adulthood [@Braganza2005]. Cases of refractory anaphylaxis are extremely rare. The incidence of refractory anaphylaxis was not investigated earlier and we estimate it to range between 3--5% of all cases of anaphylaxis (based on our clinical experience).

@@

We performed a database search in the "PubMed" database using the "RISmed"[@rismed] package for R statistical software [@R]. The search terms contained the words "refractory anaphylaxis" in articles published from 1950 till 2017. The search was conducted on 16.11.2017 and returned 151 results. All case reports of patients who experienced an anaphylactic event and did not respond to at least 600 µg of adrenaline were included in this study. Articles in English, Portuguese, French, and Spanish were included in the analysis, one article in Japanese was excluded. Additionally, case reports referenced in the included sources were added. We excluded reviews and articles not containing a case description as well as cases which documented an appropriate response to injected adrenaline. Also, cases describing mast cell activation syndrome and refractory vasoplegia without the definitive diagnosis of anaphylaxis were not included. The detailed inclusion flowchart is illustrated in Fig \@ref(fig:prismaflow).

@@

We finally included `r unique(d$Source) %>% length()` Papers reporting `r d$Author %>% length()` cases of anaphylaxis (Table \@ref(tab:demo))

@@

A major risk factor for refractory anaphylaxis is perioperative anaphylaxis. In `r (d$Situation_gr=="OP") %>% which %>% length()` (`r ((d$Situation_gr=="OP") %>% which %>% length()/length(d[,1])*100) %>% round(1)`%) cases the anaphylactic reaction occurred during a diagnostic or surgical procedure. The most common causes of refractory anaphylaxis based on the analyzed case reports were iatrogenic procedures including anesthesia drugs, aprotinin, and protamine, radiocontrast media (RCM). Indirectly concomitant treatment with protimne-containing insulins might predispose to refractory anaphylaxis as these patiens have 40 to 50 times greater risk for developing anaphylaxis perioperatively [@Stewart1984; @DrugAllergy2010]. Surprisingly, no cases of insect venom anaphylaxis were found among these reports (see Table \@ref(tab:demograph)). While the incidence of perioperative anaphylaxis has been reported to be between 1 in 10,000 -- 20,000 anesthesia procedures, it has a relatively high fatality rate of 3 -- 10% of the perioperative fatalities. [@Nag2017]

The patomechanisms which are responsible for refractory anaphylaxis are not well understood. There are however theories why extensive therapeutic efforts fail in certain cases. Figure \@ref(fig:pat) summarizes the potential factors contributing to refctorinness of anaphylaxis. These can be divided into patient dependent (e.g. concomitant coronary disease, older age or bronchial hyperresponsiveness) or therapy-dependent (e.g. suboptimal management, concomitant beta blocker use, or idopathic triggers).

@@

As methylene blue does not block nitric oxide production its side effects are minimal (nausea, vomiting, abdominal pain, fever, hemolysis, hypotension, methemoglobinemia, arrhythmias, bluish skin, urine discol- oration, and hyperbilirubinemia [@Evora2007]). It may be considered a safe option for the treatment of anaphylaxis upon intensive care conditions. No side effects were reported due to methylene blue use in the reported-cases.

@@

One case report described the use of military anti-shock trousers to perform external vasopressin on a patient who received inadvertently high doses of systemic immune therapy allergen [@Bickell1984]. Although this can be hardly available in critical care centers we want to point out this minimally invasive method as a possible alternative measure in achieving normotension.

@@

Adrenaline may fail to inhibit an anaphylactic event in patients taking beta-blockers as both heart and lung express beta-adrenergic receptors.
Glucagon, a polypeptide hormone with potent ino- and chronotropic actions was given in two cases of refractory hypotension in analyzed cases[@Zaloga1986; @Javeed1996]. Both patients receiving glucagon were undergoing refractory anaphylaxis with the concomittant use of atenolol for arterial hypertension. Both patients reacted to contrast medium and after minutes developed severe anaphylaxis not responding to repeated doses of adrenaline. An infusion of 1 mg glucagon per hour or 1 mg glucagon as a bolus relieved the refractory shock in both cases and it also successfully treated a recurrent biphasic reaction [@Javeed1996].

Adrenergic effects of glucagon are minimally influenced by beta-blockers [@Zaloga1986] as glucagon activates adenylyl cyclase directly [@Kemp2014] Therefore, it is especially useful in patients who are on long term beta-blocker treatments. Patients with concomittant beta-blockers often show bradycardia or normocardia during anaphylaxis [@Nag2017], as the reflexory tachycardia is suppressed by the beta-blockade. Authors who reported a fatal anaphylaxis case potentiated with beta blockers did not treat with glucagon but discuss this therapeutic option positively as well [@Laxenaire1984].

@@

In the most severe cases, where all pharmacologic attempts fail to alleviate symptoms of anaphylaxis clinicians turned to last resort of extracorporeal life support (ECLS) options. There were 4 reports where patients reacted extremely severe and all were operated on because of a heart condition in general anesthesia [`r which.sources(d$Source,"ECLS",d$What.helped_gr)`]. Two patients reacted to chlorhexidine, one to rocuronium and in one case the elicitor could not be established. Only one of these patients was taking a beta blocker [@Yee2016]. All these patients received multiple doses of adrenaline (total dose ranged 1 -- 15 mg), vasopressors, high volume replacement therapy, corticosteroids, and antihistaminics. Upon prolonged resuscitation procedures two of these patients were placed on cardiopulmonary bypass (CPB) and two on an extracorporeal membrane oxygenation (ECMO) device. After stabilizing of their condition, patients were weaned off from these supportive measures without neurological sequelae.

@@

We highlight therapeutic options that should be evaluated in case of anaphylaxis non-responsive to first-line therapy with adrenaline: methylene blue, vasopressors, glucagon and extracorporeal life support. We suggest a management algorithm for refractory anaphylaxis (see Fig. \@ref(fig:algo)) that is supplementary to the established protocols.

@@

Severe anaphylaxis is a critical medical condition requiring an immediate intervention. According to the international guidelines adrenaline given intramuscular (i.m.) is a rapid and relatively safe treatment stabilizing the symptoms quickly in a given patient. However, in a few patients, this intervention might not be sufficient and, concomitantly to repeating its doses, other therapeutical measures will need to be applied. Such cases are regarded as refractory anaphylaxis and are particularly seen when the elicitor of the reaction is reaching the organism intravenously. This route is associated with a high systemic load of a given allergen or other mast cell activating molecule inducing a rapid onset of mediator release like histamine, but also tryptase and chymase, which are proteases and which may in addition to histamine promote severely anaphylaxis via activation of the plasma kallikrein and to enhancement of vascular permeability [@Imamura1996].

In clinical practice, it is necessary to identify patients at risk. Such risk factors should particularly be assessed in patients undergoing medical diagnostic and therapeutic measures. The diagnosis of anaphylaxis is based on non-specific clinical presentation, most frequently with hypotension, bronchospasm and skin flushing. In cases where anaphylaxis is possible but the symptoms are not clearly apparent introduction of a point-of-case test could confirm the suspition in order to introduce immediate therapy with adrenaline. The measurement of tryptase enables to identify patients who are at risk for severe anaphylaxis (specificity of 0.93) but its sensitivity of 0.36 is insufficient [@Brown2004].

@@

@@

```{r algo, echo=F,fig.cap="The algorithm for refractory anaphylaxis management"}
knitr::include_graphics("../figures/algo.png")
```

```{r pato, echo=F,fig.cap="The patophyisiology of refractory anaphylaxis. The proposed mechanisms of anaphylaxis reaction non responding to therapy are divided into patient dependent and iatrogenic factors."}
knitr::include_graphics("../figures/pato.png")
```

@@

```{r demograph}
knitr::kable(demogr.tab,col.names = c("Elicitor","No. cases","Age","Male [%]","Perioperative [%]"),digits = 1,
             caption = "Refractory anaphylaxis cases broken down according to the elicitor of the reaction. Male - percentage of male cases to all cases ina griven group. Perioperative - fraction of anaphylaxis episode that happened either during surgery or in temporal proximity to a surgical procedure. The eliciting drugs were: cefuroxime, chlorhexidine, protamine, aprotinin, succinylocholine, gelofusine, recuronium, chemotherapy,  amoxicillin, ketokonazole, metamizole")
```

```{r wholeTab}
knitr::kable(whole.tab,
             caption = "List of papers and anaphylaxis cases included in the study, summarizing their elicitors, and effective therapy interventions. Cases where the reaction was fatal and all therapeutic options fialed were indicated with the asterix (*). Other cases resulted with full recovery of the patient. ")
```

>>>>

Division of Allergy and Immunology, Department of Dermatology, Venerology and Allergology, Charité – Universitätsmedizin Berlin, corporate member of Freie Universität Berlin, Humboldt-Universität zu Berlin, and Berlin Institute of Health, Charitéplatz 1, 10117 Berlin,

@@

# Introduction
Anaphylaxis is a potentially life-threatening, primarily mast cell-dependent reaction to most frequently food, drugs, insect venom. Although anaphylaxis remains to be a clinical diagnosis [@Kemp2014], it is highly likely if any of the following 3 criterita are fulfilled: 1) acute onset of an illness with generalized skin or mucosal involvement with flushing, hives or pruritus; 2) acute cutaneous, cardiorespiratory or gastrointestinal symptoms after exposure to a likely allegen; c) acute reduction of blood pressure (BP) after exposure to a known allergen [@Sampson2006]. The definition of anaphylaxis remains controverse [@Lieberman2015]. The pathogenesis of anaphylaxis is related to a sudden, systemic, (mostly IgE dependent) degranulation of mast cells or basophils leading to the release of multiple mediators like histamine, tryptase, chymase, platelet-activating factor, prostaglandins, and leukotrienes [@Houser2015].

By contrast, the definition of refractory anaphylaxis has not been established yet. For the purpose of this review, we define refractory anaphylaxis as anaphylaxis (meeting the criteria by NIAID/FAAN  [@Sampson2006]) unresponsive to the treatment with at least 2 doses of a minimum 300 µg adrenaline. Unresponsiveness, in this case, is defined as a lack of expected normalization of clinical symptoms (i.e. rapid increase in blood pressure, bronchodilatation, tachycardia). Previously authors suggested using the term in cases of persistent severe shock after 10 minutes of adequate resustitation [@Gouel2016].

It is important to distinguish refractory anaphylaxis (which is unresponsive to treatment) from recurrent anaphylaxis which can be defined as multiple episodes of anaphylaxis over a period of time that responded to treatment with adrenaline [@Mullins2003]. In one case report authors used the term refractory anaphylaxis to describe 6 recurrent episodes of anaphylaxis associated with the monoclonal mast cell activation syndrome, which responded well to intramuscular doses of epinephrine [@Jagdis2014] and were ultimately treated with omalizumab. We did not include such reports into our analysis and suggest to use the term “recurrent” anaphylaxis instead of “refractory” in such cases.

The incidence of anaphylaxis is increasing worldwide [@Koplin2011] and was lately estimated to be between 50 and 112 episodes per 100,000 person-years with an estimated prevalence of 0.3% to 5.1% [@TejedorAlonso2015].  It is widely believed to be an underrecognized and undertreated medical emergency [@Fineman2015]. Anaphylaxis in children is in the majority of cases food dependent and most likely less severe than episodes in adulthood [@Braganza2005]. Cases of refractory anaphylaxis are extremely rare. The incidence of refractory anaphylaxis was not investigated earlier and we estimate it to range between 3--5% of all cases of anaphylaxis (based on our clinical experience).

@@

We performed a database search in the "PubMed" database using the "RISmed"[@rismed] package for R statistical software [@R]. The search terms contained the words "refractory anaphylaxis" in articles published from 1950 till 2017. The search was conducted on 16.11.2017 and returned 151 results. All case reports of patients who experienced an anaphylactic event and did not respond to at least 600 µg of adrenaline were included in this study. Articles in English, Portuguese, French, and Spanish were included in the analysis, one article in Japanese was excluded. Additionally, case reports referenced in the included sources were added. We excluded reviews and articles not containing a case description as well as cases which documented an appropriate response to injected adrenaline. Also, cases describing mast cell activation syndrome and refractory vasoplegia without the definitive diagnosis of anaphylaxis were not included. The detailed inclusion flowchart is illustrated in Fig. \@ref(fig:prismaflow).

@@

We finally included `r unique(d$Source) %>% length()` Papers reporting `r d$Author %>% length()` cases of refractory anaphylaxis (Table \@ref(tab:demo)).

@@

A major risk factor for refractory anaphylaxis is perioperative anaphylaxis. In `r (d$Situation_gr=="OP") %>% which %>% length()` (`r ((d$Situation_gr=="OP") %>% which %>% length()/length(d[,1])*100) %>% round(1)`%) cases the anaphylactic reaction occurred during a diagnostic or surgical procedure. The most common causes of refractory anaphylaxis based on the analyzed case reports were iatrogenic procedures including anesthesia drugs, aprotinin, and protamine, radiocontrast media (RCM). Indirectly concomitant treatment with protimne-containing insulins might predispose to refractory anaphylaxis as these patiens have 40 to 50 times greater risk for developing anaphylaxis perioperatively [@Stewart1984; @DrugAllergy2010]. Surprisingly, no cases of insect venom anaphylaxis were found among these reports (Table \@ref(tab:demograph)). While the incidence of perioperative anaphylaxis has been reported to be between 1 in 10,000 -- 20,000 anesthesia procedures, it has a relatively high fatality rate of 3 -- 10% of the perioperative fatalities. [@Nag2017]

The pathomechanisms which are responsible for refractory anaphylaxis are not well understood. There are however theories why extensive therapeutic efforts fail in certain cases. Fig. \@ref(fig:pato) summarizes the potential factors contributing to refctorinness of anaphylaxis. These can be divided into patient dependent (e.g. concomitant coronary disease, older age or bronchial hyperresponsiveness) or therapy-dependent (e.g. suboptimal management, concomitant beta blocker use, or idopathic triggers).

@@

As methylene blue does not block nitric oxide production its side effects are minimal (nausea, vomiting, abdominal pain, fever, hemolysis, hypotension, methemoglobinemia, arrhythmias, bluish skin, urine discoloration, and hyperbilirubinemia [@Evora2007]). It may be considered a safe option for the treatment of anaphylaxis upon intensive care conditions. No side effects were reported due to methylene blue use in the reported cases.

@@

One case report described the use of military anti-shock trousers to perform external vasopression on a patient who received inadvertently high doses of systemic immune therapy allergen [@Bickell1984]. Although this can be hardly available in critical care centers we want to point out this minimally invasive method as a possible alternative measure in achieving normotension.

@@

Adrenaline may fail to inhibit an anaphylactic event in patients taking beta-blockers as both heart and lung express beta-adrenergic receptors.
Glucagon, a polypeptide hormone with potent ino- and chronotropic actions was given in two cases of refractory hypotension [@Zaloga1986; @Javeed1996]. Both patients receiving glucagon were undergoing refractory anaphylaxis with the concomittant use of atenolol for arterial hypertension. Both patients reacted to contrast medium and after minutes developed severe anaphylaxis not responding to repeated doses of adrenaline. An infusion of 1 mg glucagon per hour or 1 mg glucagon as a bolus relieved the refractory shock in both cases and it also successfully treated a recurrent biphasic reaction [@Javeed1996].

Adrenergic effects of glucagon are minimally influenced by beta-blockers [@Zaloga1986] as glucagon activates adenylyl cyclase directly [@Kemp2014]. Therefore, it is especially useful in patients who are on long term beta-blocker treatments. Patients with concomittant beta-blockers often show bradycardia or normocardia during anaphylaxis [@Nag2017], as the reflexory tachycardia is suppressed by the beta-blockade. Authors who reported a fatal anaphylaxis case potentiated with beta blockers did not treat with glucagon but discuss this therapeutic option positively as well [@Laxenaire1984].

@@

In the most severe cases, where all pharmacologic attempts fail to alleviate symptoms of anaphylaxis clinicians turned to last resort of extracorporeal life support (ECLS) options. There were 4 reports where patients reacted extremely severe and all were operated on because of a heart condition in general anesthesia [`r which.sources(d$Source,"ECLS",d$What.helped_gr)`]. Two patients reacted to chlorhexidine, one to rocuronium and in one case the elicitor could not be established. Only one of these patients was taking a beta blocker [@Yee2016]. All these patients received multiple doses of adrenaline (total dose ranged 1 -- 15 mg), vasopressors, high volume replacement therapy, corticosteroids, and antihistaminics. Upon prolonged resuscitation procedures two of these patients were placed on cardiopulmonary bypass (CPB) and two on an extracorporeal membrane oxygenation (ECMO) device. After stabilizing of their condition, patients were weaned off from these supportive measures without neurological sequelae. Gouel-Chéron et al. reported a fifth case of refractory anaphylaxis where the patient died on the 5th day after introducing ECLS [@Gouel].

@@

We highlight therapeutic options that should be evaluated in case of anaphylaxis non-responsive to first-line therapy with adrenaline: methylene blue, vasopressors, glucagon and extracorporeal life support. We suggest a management algorithm for refractory anaphylaxis (Fig. \@ref(fig:algo)) that is supplementary to the established protocols.

@@

Severe anaphylaxis is a critical medical condition requiring an immediate intervention. According to the international guidelines, adrenaline given intramuscular (i.m.) is a rapid and relatively safe treatment stabilizing the symptoms quickly in a given patient. However, in a few patients, this intervention might not be sufficient and, concomitantly to repeating its doses, other therapeutical measures will need to be applied. Such cases are regarded as refractory anaphylaxis and are particularly seen when the elicitor of the reaction is reaching the organism intravenously. This route is associated with a high systemic load of a given allergen or other mast cell activating molecule inducing a rapid onset of mediator release like histamine, but also tryptase and chymase, which are proteases and which may in addition to histamine promote severely anaphylaxis via activation of the plasma kallikrein and to enhancement of vascular permeability [@Imamura1996].

In clinical practice, it is necessary to identify patients at risk. Such risk factors should particularly be assessed in patients undergoing medical diagnostic and therapeutic measures. The diagnosis of anaphylaxis is based on non-specific clinical presentation, most frequently with hypotension, bronchospasm and skin flushing. In cases where anaphylaxis is possible but the symptoms are not clearly apparent introduction of a point-of-care test could confirm the suspicion in order to introduce the immediate therapy with adrenaline. The measurement of peak tryptase levels (with the cut off value of 12 µg/L) could be used for confirming the diagnosis of anaphylaxis (specificity of 0.93) but its sensitivity of 0.36 is insufficient [@Brown2004].

@@

Aditionally, based on clinical features, we will be able to predict which patients are at risk of developing severe anaphylaxis. Therefore extraordinary preventive measures could be undertaken in such patients beforehand.

@@

```{r pato, echo=F,fig.cap="The overview of factors influencing the onset of refractory anaphylaxis. "}
knitr::include_graphics("../figures/pato.png")
```

```{r algo, echo=F,fig.cap="Proposed algorithm for refractory anaphylaxis management"}
knitr::include_graphics("../figures/algo.png")
```

@@

```{r demograph}
knitr::kable(demogr.tab,col.names = c("Elicitor","No. cases","Age","Male [%]","Perioperative [%]"),digits = 1,
             caption = "Refractory anaphylaxis cases broken down according to the elicitor of the reaction. Male - percentage of male cases to all cases in a given group. Perioperative - fraction of anaphylaxis episode that happened either during surgery or in temporal proximity to a surgical procedure. The eliciting drugs were: cefuroxime, chlorhexidine, protamine, aprotinin, succinylocholine, gelofusine, recuronium, chemotherapy,  amoxicillin, ketokonazole, metamizole.")
```

```{r wholeTab}
knitr::kable(whole.tab, 
             caption = "List of reports included into this study, summarizing their elicitors, and effective therapy interventions. Cases where the reaction was fatal and all therapeutic options fialed were indicated with the asterix (*). Other cases resulted with full recovery of the patient. 'Therapy' column describes the approach which lead to a rapid recovery in a given case.  HT - arterial hypertension and DM - diabetes mellitus are reported (other concomitant diseases are not included in this table). OP - perioperative anaphylaxis.",
            col.names = c("Author","Elicitor","Age","Sex","Seting","Therapy","Conc. disease","Vaopressors"))
```
